---
title: "irt_metric_understanding"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(mirt)
library(MASS)
source("leah/my_leah_code_rewrite_functions.R")
dat <- expand.table(LSAT7)
```

fit models

```{r}
mod1 <- 
    mirt(
        dat, 1, "2PL", constrain = list(c(1, 5, 9, 13, 17)), 
        SE = TRUE
    )

mod2 <- mirt(dat, 1, "2PL", SE = TRUE)
```

calculate probs

```{r}
quads <- seq(-4, 4, by = 0.25)
probs_true_mod1 <- model_to_probs(mod1)
probs_true_mod2 <- model_to_probs(mod2)

model <- mod2
```

extract parameters and covariance matrices 

```{r}
pars_true_mod1 <- extract.mirt(mod1, "parvec")
pars_true_mod2 <- extract.mirt(mod2, "parvec")

pars_vcov_mod1 <- extract.mirt(mod1, "vcov")
pars_vcov_mod2 <- extract.mirt(mod2, "vcov")
```

draw some alternative parameters

```{r}
n_draws <- 1000
pars_draws_mod1 <- mvrnorm(n = n_draws, mu = pars_true_mod1, Sigma = pars_vcov_mod1)
pars_draws_mod2 <- mvrnorm(n = n_draws, mu = pars_true_mod2, Sigma = pars_vcov_mod2)
```

calculate probs with those alternative parameters

```{r}
probs_draws_mod1_item1 <- 
    t(apply(pars_draws_mod1, 1, function(p) 1 / (1 + exp(-p[1] * quads - p[2])))) %>% 
    as_tibble() %>% 
    set_names(paste0("th", quads))
probs_draws_mod1_item2 <- 
    t(apply(pars_draws_mod1, 1, function(p) 1 / (1 + exp(-p[1] * quads - p[3])))) %>% 
    as_tibble() %>% 
    set_names(paste0("th", quads))
probs_draws_mod1_item3 <- 
    t(apply(pars_draws_mod1, 1, function(p) 1 / (1 + exp(-p[1] * quads - p[4])))) %>% 
    as_tibble() %>% 
    set_names(paste0("th", quads))
probs_draws_mod1_item4 <- 
    t(apply(pars_draws_mod1, 1, function(p) 1 / (1 + exp(-p[1] * quads - p[5])))) %>% 
    as_tibble() %>% 
    set_names(paste0("th", quads))
probs_draws_mod1_item5 <- 
    t(apply(pars_draws_mod1, 1, function(p) 1 / (1 + exp(-p[1] * quads - p[6])))) %>% 
    as_tibble() %>% 
    set_names(paste0("th", quads))

probs_draws_mod2_item1 <- 
    t(apply(pars_draws_mod2, 1, function(p) 1 / (1 + exp(-p[1] * quads - p[2])))) %>% 
    as_tibble() %>% 
    set_names(paste0("th", quads))
probs_draws_mod2_item2 <- 
    t(apply(pars_draws_mod2, 1, function(p) 1 / (1 + exp(-p[3] * quads - p[4])))) %>% 
    as_tibble() %>% 
    set_names(paste0("th", quads))
probs_draws_mod2_item3 <- 
    t(apply(pars_draws_mod2, 1, function(p) 1 / (1 + exp(-p[5] * quads - p[6])))) %>% 
    as_tibble() %>% 
    set_names(paste0("th", quads))
probs_draws_mod2_item4 <- 
    t(apply(pars_draws_mod2, 1, function(p) 1 / (1 + exp(-p[7] * quads - p[8])))) %>% 
    as_tibble() %>% 
    set_names(paste0("th", quads))
probs_draws_mod2_item5 <- 
    t(apply(pars_draws_mod2, 1, function(p) 1 / (1 + exp(-p[9] * quads - p[10])))) %>% 
    as_tibble() %>% 
    set_names(paste0("th", quads))
```

graph the model estimate and then 2.5% and 97.5% probabilites for each model

```{r}
tibble(quads) %>% 
    mutate(
        p1 = 1 / (1 + exp(-pars_true_mod1[1] * quads - pars_true_mod1[4])),
        mod1_low = probs_draws_mod1_item3 %>% map_dbl(quantile, probs = 0.025),
        mod1_high = probs_draws_mod1_item3 %>% map_dbl(quantile, probs = 0.975),
        p2 = 1 / (1 + exp(-pars_true_mod2[5] * quads - pars_true_mod2[6])),
        mod2_low = probs_draws_mod2_item3 %>% map_dbl(quantile, probs = 0.025),
        mod2_high = probs_draws_mod2_item3 %>% map_dbl(quantile, probs = 0.975)
    ) %>% 
    gather(var, val, -quads)  %>% 
    ggplot(
        aes(x = quads, y = val, color = var, linetype = str_detect(var, "mod"))
    ) +
    geom_line() +
    facet_wrap(~ str_detect(var, "2"), ncol = 1)
```

from pars and theta to vector of probs

```{r}
mod1_pars_theta_to_prob <- function(p, theta){
    x <- c(1 / (1 + exp(-p[1] * theta - p[2])),
      1 / (1 + exp(-p[1] * theta - p[3])),
      1 / (1 + exp(-p[1] * theta - p[4])),
      1 / (1 + exp(-p[1] * theta - p[5])),
      1 / (1 + exp(-p[1] * theta - p[6])))
    names(x) <- NULL
    x
}

mod2_pars_theta_to_prob <- function(p, theta){
    x <- c(1 / (1 + exp(-p[1] * theta - p[2])),
      1 / (1 + exp(-p[3] * theta - p[4])),
      1 / (1 + exp(-p[5] * theta - p[6])),
      1 / (1 + exp(-p[7] * theta - p[8])),
      1 / (1 + exp(-p[9] * theta - p[10])))
    names(x) <- NULL
    x
}

mod1_pars_theta_to_prob(pars_true_mod1, theta = 0)

mod1_pars_theta_to_prob(pars_draws_mod1[1 , ], theta = 0)

mod2_pars_theta_to_prob(pars_true_mod2, theta = 0)

mod2_pars_theta_to_prob(pars_draws_mod2[1 , ], theta = 0)
```

optimization functions

```{r}
f1a <- function(eta, pars, targetp){
    p <- mod1_pars_theta_to_prob(pars, eta)
    sqrt(mean((p - targetp)^2))
}

f2a <- function(eta, pars, targetp){
    p <- mod2_pars_theta_to_prob(pars, eta)
    sqrt(mean((p - targetp)^2))
}

res_1a <- lapply(quads, function(t){
    targetp <- mod1_pars_theta_to_prob(p = pars_true_mod1, theta = t)
    apply(pars_draws_mod1, 1, function(pars){
        optimize(f = f1a, interval = c(-6, 6), pars = pars, targetp = targetp)
    })
})

res_2a <- lapply(quads, function(t){
    targetp <- mod2_pars_theta_to_prob(p = pars_true_mod2, theta = t)
    apply(pars_draws_mod2, 1, function(pars){
        optimize(f = f2a, interval = c(-6, 6), pars = pars, targetp = targetp)
    })
})
```

for each draw what we have is the value of theta such that we get lowest rmse with that actual model probability vector

```{r}
rmse <- function(x, y){sqrt(mean((x - y)^2))}

out <- 
    tibble(quads) %>% 
    mutate(
        df_mod1 = res_1a %>% map(bind_rows),
        df_mod2 = res_2a %>% map(bind_rows)
    )

# 1st inner row is for the first draw, which value of theta will 
# have a small rmse between prob vector under the draw and prob vector under the model
out$df_mod1[[1]]
prob_vector_model <- mod1_pars_theta_to_prob(pars_true_mod1, -4)
prob_vector_draw <- mod1_pars_theta_to_prob(pars_draws_mod1[1 , ], -3.79)
rmse(prob_vector_model, prob_vector_draw)
```

a metric is in what percent of draws do we get the order of the estimated theta perfectly correct

```{r}
out2 <- 
    out %>% 
    unnest() %>% 
    rename(
        min_mod1 = minimum, 
        obj_mod1 = objective,
        min_mod2 = minimum1, 
        obj_mod2 = objective1,
    ) %>% 
    group_by(quads) %>% 
    mutate(draw = row_number()) %>% 
    ungroup()

results <- 
    tibble(
        draw = 1:n_draws,
        mod1_spearman =
            out2 %>% split(.$draw) %>% map_dbl(~ cor(.$min_mod1, quads, method = "spearman")),
        mod2_spearman =
            out2 %>% split(.$draw) %>% map_dbl(~ cor(.$min_mod2, quads, method = "spearman"))
    )

results %>% select(-draw) %>% map_dbl(~ mean(. == 1))
```

graphing model 2

```{r}
LSATdat <- tibble("theta" = quads, 
                  "i1m" = 1 / (1 + exp(-pars_true_mod2[1] * quads - pars_true_mod2[2])),
                  "i2m" = 1 / (1 + exp(-pars_true_mod2[3] * quads - pars_true_mod2[4])),
                  "i3m" = 1 / (1 + exp(-pars_true_mod2[5] * quads - pars_true_mod2[6])),
                  "i4m" = 1 / (1 + exp(-pars_true_mod2[7] * quads - pars_true_mod2[8])),
                  "i5m" = 1 / (1 + exp(-pars_true_mod2[9] * quads - pars_true_mod2[10])),
                  "i1l" = apply(probs_draws_mod2_item1, 2, quantile, probs = .025),
                  "i2l" = apply(probs_draws_mod2_item2, 2, quantile, probs = .025),
                  "i3l" = apply(probs_draws_mod2_item3, 2, quantile, probs = .025),
                  "i4l" = apply(probs_draws_mod2_item4, 2, quantile, probs = .025),
                  "i5l" = apply(probs_draws_mod2_item5, 2, quantile, probs = .025),
                  "i1u" = apply(probs_draws_mod2_item1, 2, quantile, probs = .975),
                  "i2u" = apply(probs_draws_mod2_item2, 2, quantile, probs = .975),
                  "i3u" = apply(probs_draws_mod2_item3, 2, quantile, probs = .975),
                  "i4u" = apply(probs_draws_mod2_item4, 2, quantile, probs = .975),
                  "i5u" = apply(probs_draws_mod2_item5, 2, quantile, probs = .975))

LSATdat %>% 
    gather(type, p, -theta) %>% 
    mutate(
        item = paste("Item", substr(type, start = 2, stop = 2)),
        bound = substr(type, start = 3, stop = 3)
    ) %>% 
    ggplot(aes(theta, p, col = item, linetype = bound)) + 
    geom_path() + 
    facet_wrap(~item) + 
    scale_linetype_manual(values=c("dashed", "solid", "dashed")) + 
    theme_light() + 
    labs(x = expression(theta), y = "Probability", title = "Model 2")
```

for each draw, see how low we can get rmse to get to with theta = -4, then take the median of how low it can go. 

intuitively, if every draw was the same (there was no instability), then we could always use theta = -4 and get 0. but the more instability the more sort of tradeoffs we face.

```{r}
LSATdat2 <- 
    tibble(
        "theta" = quads,
        "M1.50" =  
            out2 %>% group_by(quads) %>% 
            summarize(o = quantile(obj_mod1, 0.5)) %>% pull(o),
        "M1.95" = 
            out2 %>% group_by(quads) %>% 
            summarize(o = quantile(obj_mod1, 0.95)) %>% pull(o),
       "M2.50" = 
           out2 %>% group_by(quads) %>% 
           summarize(o = quantile(obj_mod2, 0.5)) %>% pull(o),
       "M2.95" = 
           out2 %>% group_by(quads) %>% 
           summarize(o = quantile(obj_mod2, 0.95)) %>% pull(o),
    )

LSATdat2 %>% 
    gather(type, RMSD, -theta) %>% 
    mutate(
        model = 
               paste0(
                   substr(type, start = 2, stop = 2), "PM"),
                    quantile = substr(type, start = 3, stop = 5),
        light = (quantile == ".50" & RMSD > .02) | (quantile == ".95" & RMSD > .05)
    ) %>% 
    ggplot(aes(theta, RMSD, col = model, linetype = quantile, size = I(1))) + 
    geom_path(na.rm=TRUE) +
    theme_minimal() + 
    geom_hline(yintercept = c(.02, .05), col = I("gray"), linetype = "dotted") + 
    theme(legend.position = "bottom") + 
    lims(y = c(0, .15)) + 
    labs(x = expression(theta), y = expression(Delta)) + 
    guides(
        linetype = guide_legend(override.aes = list(color = "black", fill = "white"))
    ) +
    scale_color_grey()
```

graph test information

```{r}
LSATdat3 <- t(sapply(quads, function(t){
    a <- pars_true_mod1[1]
    b <- -pars_true_mod1[-1] / a
    info1 <- sum(catR::Ii(t, it = cbind(a, b, 0, 1))$Ii)
    a <- pars_true_mod2[c(1, 3, 5, 7, 9)]
    b <- -pars_true_mod2[c(2, 4, 6, 8, 10)] / a
    info2 <- sum(catR::Ii(t, cbind(a, b, 0, 1))$Ii)
    c(info1, info2)
}))

tibble(theta = quads, "1PM" = LSATdat3[, 1], "2PM" = LSATdat3[, 2]) %>% 
    gather(model, info, -theta) %>% 
    ggplot(aes(theta, info, col = model, size = I(1))) + geom_path() +
    scale_color_grey() + 
    theme_minimal() + 
    labs(x = expression(theta), y = "information") + theme(legend.position = "bottom")

# something was wrong so here's my double check
testinfo(mod1, 0)

tibble(
    theta = quads,
    mod1_info = quads %>% map_dbl(~ testinfo(mod1, .)),
    mod2_info = quads %>% map_dbl(~ testinfo(mod2, .)),
) %>% 
    gather(model, info, -theta) %>% 
    ggplot(aes(x = theta, y = info, color = model)) +
    geom_line()
```
